---
title: "Introduction to Predictive Modeling with Pipelines"
format: html
---

## Data Context 

The data for today is taken from [this Kaggle dataset](https://www.kaggle.com/datasets/xavya77/nhl04to18) of NHL player
statistics. The data contain statistics on every NHL player for the 
2004 through 2018 seasons assembled by Xavya. 

- Data were scraped from: <https://www.hockey-reference.com/> 
- Skater statistics were scraped from: <https://www.hockey-reference.com/leagues/NHL_2018_skaters.html>
- Hart MVP voting was scraped from:
<https://www.hockey-reference.com/awards/hart.html>

We also added in [this Tidy Tuesday dataset](https://github.com/rfordatascience/tidytuesday/blob/main/data/2024/2024-01-09/readme.md) of player height, weight, and birthday information.

This information was acquired through the [NHL API](https://github.com/Zmalski/NHL-API-Reference). 

## Loading the Data 

I've saved the `.csv` to Dropbox, so we can read it in using a URL. Run the code
provided to load the data.

```{r}
#| label: load-data-R
#| echo: true

library(tidyverse)

data_nhl <- read_csv("https://www.dropbox.com/scl/fi/rrbll4jm20fvol58nk8wf/nhl_player_stats.csv?rlkey=pbg8avxv02w8gras3mmimmifh&st=llz0bxpa&dl=0")
```

## Inspecting the Data

Take a few minutes to look over the data. 

- What is the **observational unit**? (i.e., what uniquely identifies a row?)

- What are the variables?
    * Are there any variables whose meaning is not obvious?

- What are the data types of the variables (e.g., number, character, date)

## Observations & Variables 

Each row is a player's statistics from one `season`. The variables measured
about each player are:

* identification: `first_name`, `last_name`
* physicals: `height`, `weight`, `age` (for that season)
* birth Information: `birth_date`, `birth_day`, `birth_month`, `birth_city`,
`birth_country`, `birth_state_province`
* position & team: `team`, `position_code` (right wing, left wing, center, 
or defense), `position_type` (forward or defense), and `sweater_number`.
* statistics for the season: `games_played`, `goals`, `assists`, `points`
(goals plus assists), `plusminus` (team goal differential when player is
playing), and `penalty_minutes`.

## Posing a Question

We will use these data to try to address this question: 

> What features of an NHL player are associated with them scoring more points
> (goals + assists) in a particular season?

# Steps 1 & 2: Cleaning & Visualizing the Data

Our goal here is to visualize variables we are interested in to look for 
interesting patterns, **while also** looking for issues or anomalies to fix in
the data.

## Numeric Variables

What are we looking for?

-   **Unusual observations** that we might need to *remove*.

-   Variables that are very **skewed** and might need to be *transformed*.

-   Variables that are extremely **multimodal** and might need to be *binned*.

-   Variables with values that we might want to **omit** from our study because
they are *not relevant* to the question.

### Investigating with Visuals

Use the template code provided to create visuals of each numeric variable in the
`data_nhl` dataset: 

- `height_in_inches`
- `weight_in_pounds`
- `age`
- `birth_month` 
- `games_played`
- `goals`
- `assists`
- `points`
- `plusminus`
- `penalty_minutes`

Do you find anything notable?

```{r}
#| label: ggplot-template-numerical

ggplot(data = data_nhl, 
       mapping = aes(x = points)
       ) +
  geom_histogram()
```

### Cleaning the Data

Based on your investigation, you will likely want to filter the data to 
exclude certain observations. Fill in the code below to filter the data (based
on Dr. T's slides):

*Replace the `___` with the relational statement you want (e.g., `> 50`)!*

```{r}
#| label: data-cleaning-numerical

data_nhl <- data_nhl |>
  filter(
    games_played ___,
    points ___,
    age ___,
    penalty_minutes ___
  )
```

## Categorical variables

What are we looking for?

-   Categories that should be **combined** to make a few larger ones.
-   Variables that should have an `"Other"` category to combine the rarest
categories.
-  Variables with **too many categories** to reasonably include them in our
analysis.
- Variables that are **subsets** of each other, so we might want to include only
one or the other.
- Anything that suggests **issues with the data** to fix.


## Investigating with Visuals

Use the template code provided to create visuals of each numeric variable in the
`data_nhl` dataset: 

- `season`
- `team`
- `position_code`,
- `position_type`
- `birth_city`
- `birth_country`
- `birth_state_province`

Do you find anything notable?

```{r}
#| label: ggplot-template-categorical

ggplot(data_nhl, 
       mapping = aes(x = position_code)) +
  geom_bar()
```

## Addressing Unusual Observations

These visualizations should reveal two issues:

1. For some reason, the entries for the `2009` season are listed twice.  We will
simply delete the duplicates.

2. The `birth_city`, `birth_state_province`, and `birth_country` variables have
far too many different values to reasonable use.  

    - We will use the `country` variable only, and we will limit its values to 
    `"Canada"`, `"United States"`, or `"Other"`.

```{r}
#| label: data-cleaning-categorical

data_nhl <- data_nhl |>
  distinct() |>
  mutate(
    can_or_usa = fct_lump()
  ) 
```

### Check in

Make sure your re-visualize your variables after making the cleaning changes, to ensure that this accomplished what you hoped!

```{r}
#| label: cleaning-check

```

# Step 3: Choosing Candidate Predictors

We will answer this question with a *predictive model*, that uses information in
some *predictor* variables to create a process for guessing the value of the
*target* variable (`points`).

To decide which variables are worth including in our eventual model, we need to
understand how they associate with the target variable of `points`.

## Numeric Variables

Consider the *numeric* variables in this dataset.  Which ones might be
reasonable to use in our model?

Use the code provided to visualize the relationship between `points` and a few
of these potential predictors. Decide which variable(s) seem to have the 
strongest relationship with our target variable (`points`). 

```{r}
#| label: ggplot-template-two-numerical

ggplot(data = data_nhl, 
       mapping = aes(x = age, y = points)
       ) +
  geom_point() +
  geom_smooth()
```

## Categorical variables

Consider the *categorical* variables in this dataset.  Which ones might be
reasonable to use in our model?

Use the code provided to visualize the relationship between `points` and a few
of these potential predictors. Decide which variable(s) seem to have the 
strongest relationship with our target variable (`points`). 

```{r}
#| label: ggplot-template-numerical-categorical

library(ggridges)

ggplot(data = data_nhl, 
       mapping = aes(x = points, y = position_type)
       ) +
  geom_density_ridges()
```

## Multiple Variables

Sometimes, the way one variable associates with `points` might be different
depending on another variables.

It is believed that players who are born earlier in the month are more likely to
become professional athletes, because they are on the older end of the league
for their age growing up.  

It may be that this trend only applies in some countries, since not all
countries have the same youth sports system. Let's see!

```{r}
#| label: points-by-birth-faceted-by-country

dat_nhl |>
  mutate(birth_month = as.factor(birth_month)) |> 
  ggplot(data = data_nhl, 
         mapping = aes(x = points, y = birth_month)
         ) +
  geom_density_ridges() +
  facet_wrap(~can_or_usa)
```


### Make predictor recipes

Now, we are ready to prepare our **recipes**: the different combinations of predictors that we think could produce a good model.

We'll try two recipes today:

* One with *all* the possible predictors

* One with only the three "best" predictors that you choose.

```{r}
rec_all <- recipe(points ~ height + weight + age + can_or_usa + birth_month + position_code + season + team + penalty_minutes, data = dat_nhl)

rec_best <- recipe(points ~ position_type + season + team , data = dat_nhl)
```




## 3. Establish candidate workflows


### Candidate model specifications

The other decision we must make - besides which *predictors* to include - is which *model types* to try.

In this analysis, we will try *Linear Regression* and a *Decision Tree*.  The Linear Regression has no additional decisions to make, but for the Decision Tree, we need to choose **hyperparameters**. 

:::{.callout-warning}

There are many, many, many different model types - most with infinite hyperparameter options - that we can use for any given modeling problem.

There is no magic formula for deciding which ones to try.  As you get more experience with machine learning, you'll develop more instincts for what range of options to try.

:::

```{r}
lr_spec <- linear_reg() |>
  set_mode("regression")

dt_spec <- decision_tree(min_n = 10) |>
  set_mode("regression") 
```




:::

### Pre-processing options

Recall that we noticed a few interesting things in our exploratory data analysis:

* `penalty_minutes` is very right-skewed 

* Birth country seems to impact how birth month impacts the target. 

We will add a *tranformation* and an *interaction* to our workflow.

```{r}
rec_all <- rec_all |>
  step_log(penalty_minutes) |>
  step_dummy(can_or_usa, position_code) |>
  step_interact(~ birth_month*contains("can_or_usa"))

rec_best <- rec_best |>
  step_log(penalty_minutes) |>
  step_dummy(can_or_usa, position_code) |>
  step_interact( ~ birth_month*contains("can_or_usa"))
```








### Data + model combinations

Finally, we will try all our different **model type** options with all our different **recipe and preprocessing** choices.

Each of these candidate options is called a **workflow** or a **pipeline**.

```{r}
wflow_lr_all <- workflow() |>
  add_model(lr_spec) |>
  add_recipe(rec_all)

wflow_lr_best <- workflow() |>
  add_model(lr_spec) |>
  add_recipe(rec_best)

wflow_dt_all <- workflow() |>
  add_model(dt_spec) |>
  add_recipe(rec_all)

wflow_dt_best <- workflow() |>
  add_model(dt_spec) |>
  add_recipe(rec_best)
```






## 4. Assess workflows

Whew!  

All this work, and we haven't even fit a model yet!  Everything we have done to this point was *setup*:  We were only establishing our plans for a few different workflows to try.

Now, how will we compare the four options and see which one is most successful at predicting `points`?  

The process goes like this:

1. Randomly set aside 20% of our data rows as a *test set*.  The rest is called the *training set*.

2. Fit all of the workflows on the *training set*.

3. See how well each fitted workflow does at predicting on the *test set*.


:::{.callout-warning}

In a real machine learning setting, we would probably use **cross-validation** to assess our models.   This essentially means we create *many different* test/training splits and then average assessment metrics.

For the sake of time, we won't do that today; but both `tidymodels` and `scikit-learn` have great shortcuts to perform this kind of assessement!

:::

### Split into test/training set

```{r}
set.seed(2244)

split <- initial_split(dat_nhl, prop = 0.8)

dat_nhl_train <- training(split)
dat_nhl_test <- testing(split)

```



### Fit models on training data

```{r}
wflow_lr_all_fit <- wflow_lr_all |> fit(dat_nhl_train)

wflow_lr_best_fit <- wflow_lr_best |> fit(dat_nhl_train)

wflow_dt_all_fit <- wflow_dt_all |> fit(dat_nhl_train)

wflow_dt_best_fit <- wflow_dt_best |> fit(dat_nhl_train)
```




### Assess models on test data

Finally, we will use these fitted models to make predictions on the test data, and we'll see how close these predictions were to the true observed values for `points`

:::{.callout-note}

In this analysis, we are using *root mean squared error* as our **metric** for the test data.  That is, we take the squared difference between the *predicted* number of points a player will get and the *actual* number of points they get - then we add those up and average them out.

There are many metrics one can use to measure "how close" a prediction is to a true observed value.  Deciding which metric captures your goals in the modeling problem is a **big decision** for a researcher!

:::

```{r}

# Generate predictions
pred_lr_all <- predict(wflow_lr_all_fit, dat_nhl_test) |> bind_cols(dat_nhl_test)

pred_lr_best <- predict(wflow_lr_best_fit, dat_nhl_test) |> bind_cols(dat_nhl_test)

pred_dt_all <- predict(wflow_dt_all_fit, dat_nhl_test) |> bind_cols(dat_nhl_test)

pred_dt_best <- predict(wflow_dt_best_fit, dat_nhl_test) |> bind_cols(dat_nhl_test)


# Compute RMSE metric

rmse(pred_lr_all, truth = points, estimate = .pred)
rmse(pred_lr_best, truth = points, estimate = .pred)
rmse(pred_dt_all, truth = points, estimate = .pred)
rmse(pred_dt_best, truth = points, estimate = .pred)

```





## 5. Deliver results

### Justify your final model

Of our four candidate workflows, which was best?  By how much?

### Fit your final model

We used out test/training split to **assess our candidates**.  However, once we've chosen a winner, we want to make sure we use *all our data* for our final conclusions!

So, we have one last model fitting task:

```{r}
final_model <- wflow_dt_all |>
  fit(dat_nhl)
```

### python




### Interpret your final model

Sometimes, the goal of your machine learning process was simply to produce this model - perhaps you will then use it on next year's NHL data to guess which players will score the most!

Other times, the goal is more about **interpretation:** you want to tell stories about the patterns in the data.

Since our best final model was a *decision tree*, we'll make a visual of the results:

```{r}
library(rpart.plot)

# Pull out the parsnip model object
tree_fit <- extract_fit_parsnip(final_model)

rpart.plot(tree_fit$fit)
```



